---
title: "Tests of association"
subtitle: "Chi-square tests for independence and association in two-way tables"
format: 
  revealjs:
    logo: img/poly-logo-2.jpg
    footer: "STAT218"
    smaller: true
    tbl-colwidths: [30, 30, 40, 30]
    mermaid:
      theme: neutral
execute: 
  echo: false
  warning: false
  message: false
html-math-method:
  method: mathjax
  url: "https://cdn.jsdelivr.net/npm/mathjax@3/es5/tex-mml-chtml.js"
---

```{r setup}
library(tidyverse)
library(oibiostat)
library(pander)

red.grad <- colorRampPalette(c('#ff6459', '#4d0500'))
add.totals <- function(tbl){
  rbind(tbl, total = colSums(tbl)) |>
  cbind(total = c(rowSums(tbl), sum(tbl)))
}
asthma <- matrix(data = c(49, 781, 30, 769), 
               nrow = 2, 
               byrow = T, 
               dimnames = list(sex = c('female', 'male'),
                               asthma = c('asthma', 'no asthma'))) |>
as.data.frame() |>
rownames_to_column('sex') |>
pivot_longer(-sex, names_to = 'asthma', values_to = 'n') |>
  group_by(sex, asthma) |>
  sample_n(size = n, replace = T) |>
  mutate(subj.id = row_number()) |>
  select(subj.id, sex, asthma)
```

## Today's agenda

1. [lecture] tests of association in contingency tables
2. [lab] $\chi^2$ tests and residual analysis in R

## Do asthma rates differ by sex?

::: {.columns}

::: {.column width="45%"}
From a subsample of NHANES data:

```{r}
asthma.tbl <- table(asthma$sex, asthma$asthma)
asthma.tbl |> pander()
```

Inference for the difference in proportions:

```{r, echo = T}
table(asthma$sex, asthma$asthma) |>
  prop.test(alternative = 'two.sided', 
            conf.level = 0.95,
            correct = F)
```
:::

::: {.column width="55%"}
Interpretation:

> There is moderate evidence that asthma prevalence differs between men and women (*Z* = 2.108, *p* = 0.0436). With 95% confidence, the difference in prevalence (F - M) is estimated to be between 0.07% and 4.22%, with a point estimate of 2.15%.

This inference relies on a specific measure of association (difference in prevalence) that we can't always estimate.

*Could we test for association between sex and asthma without relying on a specific measure?*
:::

:::

## Association and independence

::: {.columns}

::: {.column width="59%"}
Consider the hypotheses:

$$
\begin{cases}
H_0: &\text{asthma}\perp\text{sex} \; &(\text{independence})\\
H_A: &\neg(\text{asthma}\perp\text{sex}) \; &(\text{association})
\end{cases}
$$

Consider also the proportions:
```{r}
data.frame(row = c(2, 2, 3), col = c(2, 3, 2)) |>
  as.matrix() |>
  emphasize.strong.cells()

asthma.tbl |>
  prop.table() |>
  add.totals() |>
  pander()
```

- $p_{ij}$: proportion of observations in cell $ij$ 
- $p_i$, $p_j$: marginal proportions in row $i$ or column $j$


:::

::: {.column width="41%"}

If sex and asthma are independent:

$$
p_{ij} \approx p_i \times p_j
$$

For example, we'd expect: 

$$
0.4721 \approx 0.4905 \times 0.9515
$$

In other words:

- 49% of respondents are men
- 95% of respondents don't have asthma
- so roughly 49% of 95% would be men without asthma


:::

:::

## Basis for a test: expected counts

Expected proportions translate directly to expected counts:
$$p_{ij} = p_i \times p_j 
\quad\Longleftrightarrow\quad n_{ij} = \frac{n_{i\cdot} \times n_{\cdot j}}{n}$$

::: {.columns}

::: {.column width="40%"}
Actual counts:

&nbsp; | O1 | O2 | total
---+---+---+---
**G1** | $n_{11}$ | $n_{12}$ | $\color{red}{n_{1\cdot}}$
**G2** | $n_{21}$ | $n_{22}$ | $\color{orange}{n_{2\cdot}}$
**total** | $\color{blue}{n_{\cdot 1}}$ | $\color{green}{n_{\cdot 2}}$ | $n$
:::

::: {.column width="60%"}
Expected counts under independence:

&nbsp; | O1 | O2 | total
---+---+---+---
**G1** | $\hat{n}_{11} = \frac{\color{red}{n_{1\cdot}} \color{black}{\times} \color{blue}{n_{\cdot 1}}}{n}$ | $\hat{n}_{12} = \frac{\color{red}{n_{1\cdot}} \color{black}{\times} \color{green}{n_{\cdot 2}}}{n}$ | $\color{red}{n_{1\cdot}}$
**G2** | $\hat{n}_{21} = \frac{\color{orange}{n_{2\cdot}} \color{black}{\times} \color{blue}{n_{\cdot 1}}}{n}$ | $\hat{n}_{22} = \frac{\color{orange}{n_{2\cdot}} \color{black}{\times} \color{green}{n_{\cdot 2}}}{n}$ | $\color{orange}{n_{2\cdot}}$
**total** | $\color{blue}{n_{\cdot 1}}$ | $\color{green}{n_{\cdot 2}}$ | $n$

:::

:::
Idea for a test of independence: 

- reject $H_0$ if actual and expected counts differ enough across the table
- *i.e.*, reject $H_0$ when $n_{ij} - \hat{n}_{ij}$ is large across $i, j$

## Computing expected counts

::: {.columns}

::: {.column width="40%"}
Actual counts:

&nbsp; | O1 | O2 | total
---+---+---+---
**G1** | $n_{11}$ | $n_{12}$ | $\color{red}{n_{1\cdot}}$
**G2** | $n_{21}$ | $n_{22}$ | $\color{orange}{n_{2\cdot}}$
**total** | $\color{blue}{n_{\cdot 1}}$ | $\color{green}{n_{\cdot 2}}$ | $n$
:::

::: {.column width="60%"}
Expected counts under independence:

&nbsp; | O1 | O2 | total
---+---+---+---
**G1** | $\hat{n}_{11} = \frac{\color{red}{n_{1\cdot}} \color{black}{\times} \color{blue}{n_{\cdot 1}}}{n}$ | $\hat{n}_{12} = \frac{\color{red}{n_{1\cdot}} \color{black}{\times} \color{green}{n_{\cdot 2}}}{n}$ | $\color{red}{n_{1\cdot}}$
**G2** | $\hat{n}_{21} = \frac{\color{orange}{n_{2\cdot}} \color{black}{\times} \color{blue}{n_{\cdot 1}}}{n}$ | $\hat{n}_{22} = \frac{\color{orange}{n_{2\cdot}} \color{black}{\times} \color{green}{n_{\cdot 2}}}{n}$ | $\color{orange}{n_{2\cdot}}$
**total** | $\color{blue}{n_{\cdot 1}}$ | $\color{green}{n_{\cdot 2}}$ | $n$

:::

:::

For the asthma example:

::: {.columns}

::: {.column}
```{r}
asthma.tbl |> add.totals() |> pander(caption = 'Actual')
```
:::

::: {.column}
```{r}
rslt <- chisq.test(asthma.tbl, correct = F)
rslt$expected |> 
  add.totals() |>
  pander(caption = 'Expected')
```

:::

:::


## The chi-square ($\chi^2$) statistic

A measure of the amount by which actual counts differ from expected counts under independence is the **chi** (pronounced /ˈkaɪ ) **square statistic**:

$$
\chi^2 = \sum_{ij} \frac{\left(n_{ij} - \hat{n}_{ij}\right)^2}{\hat{n}_{ij}} 
\qquad\left(\sum_\text{all cells} \frac{(\text{observed} - \text{expected})^2}{\text{expected}}\right)
$$

::: {.columns}

::: {.column}
Cell-wise calculation:

------------------------------------------------------------------------------
   &nbsp;     asthma                           no asthma 
------------ -------------------------------- --------------------------------
 **female**   $\frac{(49 - 40.25)^2}{40.25}$  $\frac{(781 - 789.7)^2}{789.7}$  

  **male**    $\frac{(30 - 38.75)^2}{38.75}$  $\frac{(769 - 760.3)^2}{760.3}$   
------------------------------------------------------------------------------
:::

::: {.column}
Result:
```{r}
rslt$residuals^2 |> pander()
```
:::

:::

Chi-square statistic: 
$$
\chi^2 
= `r rslt$residuals^2 |> as.numeric() |> round(4) |> str_flatten(collapse = ' + ')` 
= `r rslt$statistic |> round(4)`
$$

## Sampling distribution for $\chi^2$

::: {.columns}

::: {.column}
Under $H_0$, the $\chi^2$ statistic has a sampling distribution that can be approximated by a $\chi^2_1$ model.

- subscript indicates degrees of freedom parameter

The model assumes no expected counts are too small.

- rule of thumb: at least 10 ($\hat{n}_{ij} \geq 10$)
- consequences: if $\hat{n}_{ij}$ are too small, the statistic is inflated relative to the model, leading to a higher type I error rate
:::

::: {.column}

```{r, fig.width = 10, fig.height = 6.5}
reds <- red.grad(5)
par(mar = c(4, 4, 2, 1), cex = 2.5)
k <- 1
curve(sqrt(dchisq(x, k)), from = 0, to = 15, n = 200,
      xlab = expr(paste(chi^2, ' statistic')),
      ylab = '',
      main = expr(paste(chi[df]^2, ' model')),
      xaxt = 'n', yaxt = 'n', axes = F,
      col = reds[1], lwd = 2)
axis(1, at = seq(0, 15, by = 2))
title(ylab = 'sampling frequency', line = 1)

for(k in 2:5){
curve(sqrt(dchisq(x, k)), from = 0, to = 15, n = 200,
      xlab = '', ylab = '', main = '',
      xaxt = 'n', yaxt = 'n',
      col = reds[k], add = T, lwd = 2)
}

legend('topright', 
       legend = (paste("df = ", 1:5, sep = '')),
       col = reds, lwd = 2, cex = 0.75)
```

:::

:::


## Computing $p$ values


::: {.columns}

::: {.column width="60%"}

$$
\begin{cases}
H_0: &\text{asthma}\perp\text{sex} \; &(\text{independence})\\
H_A: &\neg(\text{asthma}\perp\text{sex}) \; &(\text{association})
\end{cases}
$$

To determine the test outcome, find the $p$-value: 

$$
P(\chi^2_1 > \chi^2_\text{obs}) = P(\chi^2_1 > 4.074) = 0.0435
$$

So if asthma and sex were independent, only 4% of random samples would produce a table that deviates from expected counts by more than what we observed.

:::

::: {.column width="40%"}
```{r, fig.width=6, fig.height=5}
par(mar = c(4, 4, 3, 1),
    cex = 1.75)
x <- seq(0.1, 10, length = 1000)
y <- sqrt(dchisq(x, df = 1))
chistat <- 4.074
curve(sqrt(dchisq(x, df = 1)), from = 0.1, to = 10, n = 500, lwd = 2,
xlab = expression(chi^2), ylab = '', main = expr(paste(chi^2, ' model')),
yaxt = 'n', axes = F)
axis(side = 1, at = (0:10))
title(ylab = 'frequency', line = 0.5)
polygon(c(chistat, x[x>chistat], max(x)), c(0, y[x>chistat] - 0.005, 0), col="red", border = NA)
abline(v = 4.074, lty = 2, col = 'red', lwd = 2)
legend(x = 'topright', 
       legend = c('4.35% of samples', '95.65% of samples'), 
       fill = c('red', 'white'))
```

```{r, echo = T}
pchisq(4.074, df = 1, lower.tail = F)
```

:::

:::

## Implementation in R

The R implementation is `chisq.test(...)`.

- input: contingency table
- no constraints on row/column arrangement

::: {.columns}

::: {.column}
```{r, echo = T}
# construct table and pass to chisq.test
table(asthma$sex, asthma$asthma) |> 
  chisq.test(correct = F)
```
:::

::: {.column}
> The data provide moderate evidence that asthma prevalence is associated with sex ($\chi^2$ = 4.074 on 1 degree of freedom, *p* = 0.0435).
:::

:::

## Residuals in $\chi^2$ tests

::: {.columns}

::: {.column}
The **residual** for each cell is defined as a standardized difference between the observed and expected count:

$$r_{ij} = \frac{n_{ij} - \hat{n}_{ij}}{\sqrt{\hat{n}_{ij}}} $$

Examining residuals can indicate the source(s) of an inferred association. 

- $r_{ij} > 0$: observation exceeds expectation
- $r_{ij} < 0$: observation is under expectation
- large $|r_{ij}|$ explain the association

:::

::: {.column}
```{r, echo = T, eval = F}
# store test result
rslt <- chisq.test(asthma.tbl, correct = F)

# examine residuals
rslt$residuals
```
```{r}
rslt <- chisq.test(asthma.tbl, correct = F)
rslt$residuals |> pander()
```

Look for the largest residuals:

> Asthma prevalence is higher-than-expected among women and lower-than-expected among men.
:::

:::

## Continuity correction

The $\chi^2$ test for independence is typically applied with Yates' continuity correction.

::: {.columns}

::: {.column}
This consists in using a modified version of the test statistic:

$$
\chi^2_\text{Yates} = \sum_{ij} \frac{\left(|n_{ij} - \hat{n}_{ij}| - 0.5\right)^2}{\hat{n}_{ij}}
$$

- every other detail of the test is the same
- doesn't change expected counts
- residuals are still computed as $\frac{n_{ij} - \hat{n}_{ij}}{\sqrt{\hat{n}_{ij}}}$
:::

::: {.column}
Implementation:
```{r, echo = T}
# construct table and pass to chisq.test
table(asthma$sex, asthma$asthma) |> 
  chisq.test(correct = T)
```

Note the larger $p$-value -- this changes the conclusion!
:::

:::

## Spot any similarities?

Compare the $\chi^2$ test with inference on the difference in proportions.

::: {.columns}

::: {.column}
```{r, echo = T}
# chi square test
table(asthma$sex, asthma$asthma) |> 
  chisq.test(correct = T)
```

```{r, echo = T}
# difference in proportions
table(asthma$sex, asthma$asthma) |> 
  prop.test(alternative = 'two.sided', 
            conf.level = 0.95, 
            correct = T)
```
:::

::: {.column}
- the tests are identical!

- the difference in proportions $\hat{p}_F - \hat{p}_M$ is one specific measure of association

- next time we'll learn about other measures, which also have the same inference attached

:::

:::


## Extending to $I \times J$ tables

::: {.columns}

::: {.column}
FAMuSS data:
```{r}
data(famuss)
tbl <- xtabs(~ race + actn3.r577x, data = famuss) 
tbl |> add.totals() |> pander()
```
:::

::: {.column}
Expected counts:
```{r}
chisq.test(tbl)$expected |> round(2) |> add.totals() |> pander()
```
:::

:::

- expected counts and chi-square statistic are calculated exactly the same way
- degrees of freedom are now $(I - 1)\times(J - 1)$
- appropriate provided all $\hat{n}_{ij} > 1$ and most (~80%) $\hat{n}_{ij} \geq 5$

## Extending to $I\times J$ tables

In detail:

---------------------------------------------------------------------------------------------------------------------
     &nbsp;       CC                                CT                                TT  
---------------- --------------------------------  --------------------------------  --------------------------------
 **African Am**   $\frac{(16 - 7.85)^2}{7.85}$     $\frac{(6 - 11.84)^2}{11.84}$     $\frac{(5 - 7.306)^2}{7.306}$  

   **Asian**      $\frac{(21 - 15.99)^2}{15.99}$    $\frac{(18 - 24.13)^2}{24.13}$    $\frac{(16 -14.88)^2}{14.88}$  

 **Caucasian**    $\frac{(125 - 135.8)^2}{135.8}$   $\frac{(216 - 204.9)^2}{204.9}$   $\frac{(126 - 126.4)^2}{126.4}$ 

  **Hispanic**     $\frac{(4 - 6.687)^2}{6.687}$    $\frac{(10 - 10.09)^2}{10.09}$     $\frac{(9 - 6.224)^2}{6.224}$  

   **Other**       $\frac{(7 - 6.687)^2}{6.687}$    $\frac{(11 - 10.09)^2}{10.09}$     $\frac{(5 - 6.224)^2}{6.224}$  
---------------------------------------------------------------------------------------------------------------------

Then:

$$\begin{cases} &\chi^2 = \sum \text{all cells above} = 19.4 \\
&P(\chi^2_{8} > 19.4) = 0.01286 \end{cases}
\quad\Longrightarrow\quad \text{reject hypothesis of no association}$$

## Inference for $I\times J$ tables in R

::: {.columns}

::: {.column}
The implementation is the same as for a $2\times 2$ table:
```{r, echo = T}
# construct table and pass to chisq.test
table(famuss$race, famuss$actn3.r577x) |>
  chisq.test()
```
:::

::: {.column}
> The data provide evidence of an association between race and genotype ($\chi^2$ = 19.4 on 8 degrees of freedom, *p = 0.01286*).

:::

:::

*Which genotype/race combinations are contributing most to this inferred association?*

## Residual analysis

::: {.columns}

::: {.column}
```{r, echo = T, eval = F}
# store result of test; display residuals
rslt <- chisq.test(tbl)
rslt$residuals
```
```{r}
# store result of test; display residuals
rslt <- chisq.test(tbl)
rslt$residuals |> pander()
```

:::

::: {.column}
Again look for the largest absolute residuals to explain inferred association.

> African American and Asian populations have higher CC and lower CT frequencies than would be expected if genotype were independent of race.
:::

:::


