---
title: "R Lab 1: Welcome to Posit Cloud!"
author:
  - "Team Captain: NAME HERE"
  - "Facilitator: NAME HERE"
  - "Recorder: NAME HERE"
  - "Resource Manager: NAME HERE"
date: "DATE HERE"
format: pdf
editor: visual
execute:
  echo: true
---

## Quarto Text Formatting

This is a Quarto document. Quarto is a simple formatting syntax for authoring HTML, PDF, and Microsoft Word documents.

The document may look a bit different that you are used to, but we will break it down!

Section headers are defined with hashtags (#), where one \# is the largest header, two \## is the second largest and so on.

You can make words boldface with two \*\* (like **so**).

You can make words italicized with one \* (like *this*).

You can make lists like this:

-   item 1
-   item 2
-   item 3

(Notice you need to have a blank line before the first item of your list!)

<!-- This is a Quarto comment. It WILL NOT show in your knitted document. -->

```{=html}
<!-- Comments are great ways to leave yourself notes inside your Quarto
document that no one can see!  -->
```
## R Code

You can differentiate the R code within an Quarto file from the body of the document, from the three tickmarks (\`\`\`) they start and end with.

Here is an example of an R code chunk:

```{r setup}
#| include: false

# This is an R comment. R knows this IS NOT code because it has a # in front! 

# Code to load in packages we need
library(tidyverse)

```

Notice that after the \`\`\` you see a curly brace with an "r" in it ({r}), this indicates to RMarkdown that the code inside of that code chunk is R code (not Python code or another type of code).

Also, the "setup" after the "r" is the code chunk's name. It is good practice to name your code chunks based on the action they perform. It makes it easier to navigate to them in the navigation pane.

1.  **What does the library() function do?**

    The reporter should type the answer here.

2.  **What is the tidyverse?**

    The reporter should type the answer here.

## Rendering

When you click the **Render** button, a document will be generated that includes both contents as well as the output of any embedded R code chunks within the document.

3.  **Why do you not see the above code chunk (named `setup`) when you render the document?**

## Including Code Output

You can include code output in your rendered document:

```{r hearing-loss}
head(mpg)
```

4.  **What do you think the `head()` function does?** *(Hint: What kind of output does it give you?)*

## Loading Data into `R`

You can load data files into R using functions from different packages. You will learn that the function you use depends on what format the data are in. Today, we will work with data that are stored in a Comma Separated Values file (a CSV). For this format of data, we will use the `read_csv()` function to read in our data.

The data are located in the *data* folder of your project. To load in the data, we need to specify the "path" to the data, or where R needs to look for the data. I'll walk you through what the code below is doing.

-   `arson_data` is the name of the object we are storing the data set into. This is the name R knows the data set by.
-   The `<-` "assigns" the output of the `read_csv()` function into the `arson` object
-   `read_csv()` is a function that reads in our csv data file
-   `"data"` is the name of the sub-folder the data live in (look in the bottom right of your RStudio IDE)
-   `"arson.csv"` is the name of the data set. This is the name Posit Cloud's data sub-folder knows the data set by.

```{r read-data}
#| include: false

arson_data <- read_csv("data/arson.csv")
```

5.  **Write code to preview the observed arson data set.**

    The reporter should type the answer here.

```{r arson-data}
# Your code for question 5 goes below this line!

```

6.  **What are the observations in this data set?**

7.  **How many observations are there in the data set?** *Hint: you can use code to find this or look in the top right environment.*

    The reporter should type the answer here.

8.  **What are the variables in this data set? What are their data types?**

    The reporter should type the answer here.

9.  **How many variables are there in the data set?** *Hint: you can use code to find this or look in the top right environment.*

    The reporter should type the answer here.

## Including Plots

You can also embed plots in the rendered document. Here is an example of a plot:

```{r plot}
#| echo: false 

ggplot(data = arson_data,
       aes(x = Outcome,
           fill = Outcome)
       ) + 
  geom_bar(stat = "count") +
  labs(title = "Arson Suspect Responses",
       x = "Outcome",
       y = "Number of Questions"
       )
```

8.  **What do you think the `#| echo: false` option does in the above code chunk?**

    The reporter should type the answer here.
