---
title: "Your Title Goes Here! Make it fun and related to your data set!"
subtitle: "Stat 218: Final Project"
author:
  - "Team Captain: NAME HERE"
  - "Facilitator: NAME HERE"
  - "Recorder: NAME HERE"
  - "Resource Manager: NAME HERE"
date: "DATE HERE"
format: pdf
editor: visual
execute:
  echo: true
  error: true
  message: false
  warning: false
---

<!-- Note, anything in green is a comment and will not show when your document is rendered. Please type your material below the green comments. -->

```{r setup}
#| include: false
# The code to load in the necessary packages goes here!
# You will need: tidyverse, mosaic, and moderndive
```

+ DATA SET NAME: <copy paste link to data set here>

```{r read-data}
## This is where you will read in your data! 
# I suggest copying and pasting code from a prior lab. 
# You will need to change: 
# 1. the name of the dataset (you get to choose the name)
# 2. directions for how to get to the data (is it in a data folder? is it in the main folder?)

```

```{r clean-data}
## Replace the DATA NAME with your data set name after reading in the data. This will make sure the names in your data set are all "clean" -- no spaces, all lowercase, etc. 
DATA NAME <- DATA NAME |> 
  janitor::clean_names()
```

```{r print-data}
## Print the top 6 rows of your data set
```

## Data Description

<!-- In this section you are to describe your data set! -->

<!-- This section has three tasks:
1. Describe the context of your data set in your own words! Make sure to tell me where the data set came from, what an observational unit is, and what variables / information is stored in the data set.
2. Tell me why your group picked this data set.
3. Describe the two quantitative variables you are interested in. Which is the the explanatory variable and which is the response? What do you expect the relationship to be?
-->

Your data set description goes here. This should be a paragraph and read like a report.


## Data Visualization
<!-- In this section you create three visualizations of your data -->

### Histogram of RESPONSE VARIABLE

```{r histogram-1}
## Code to make your first histogram goes here! 
## Don't forget to choose nice binwidths (be dramatic) and axis titles!
```

<!-- Here is where you describe what you see in the histogram. Your description must include the following components: center, spread, shape, outliers. Your group may choose to additionally run the favstats() function to get summary statistics for this variable. -->

Describe the histogram of the response variable here.


### Histogram of EXPLANATORY VARIABLE

```{r histogram-2}
## Code to make your second histogram goes here! 
## Don't forget to choose nice binwidths and an axis title!


```

<!-- Here is where you describe what you see in the histogram. Your description must include the following components: center, spread, shape, outliers.  Your group may choose to additionally run the favstats() function to get summary statistics for this variable.  -->

Describe the histogram of the explanatory variable here.

### Scatterplot of RESPONSE VARIABLE and EXPLANATORY VARIABLE

```{r scatterplot}
## Code to make your scatterplot goes here! 
## Don't forget to choose nice axis labels!
```

<!-- Here is where you describe what you see in the scatterplot. Your description must include the following components: direction, shape, strength, outliers -->

Describe scatterplot here.

## Linear Relationship Between <INSERT NAME OF RESPONSE VARIABLE> and <INSERT NAME OF EXPLANATORY VARIABLE>

### Model Conditions
<!-- Here is where you describe the four conditions (LINE) for linear regression.  -->
<!-- State what the condition is and if you believe it is / is not violated IN THE CONTEXT OF YOUR DATA! -->

Put conditions here. This should be in a paragraph form with complete sentences.

### Coefficient Estimates

```{r lm}
## Code to fit a linear regression (note this, is different from adding the line to the scatterplot).
```

```{r coefs}
## Code to get coefficient table from linear regression
```

Make sure to type the equation (in context of your data) and interpret the coefficients in context of your data set here.

### Inference for the Slope

<!-- Here is where you describe the hypothesis test and results for whether the slope is 0.  -->

Put slope hypothesis test paragraph here. Make sure to include all parts necessary for a hypothesis test from setup to conclusion. Make sure to use context of your data.

<!-- Here is where you describe the confidence interval for the true slope. -->

Put interpretation of confidence interval for the true slope here. Make sure to use context of your data.
