---
title: 'Lab 8: Soil Moisture & Early Growth of Strawberries'
subtitle: "Chapter 8: Comparing a Numerical Variable Across More Than Two Groups"
author:
  - "Team Captain: NAME HERE"
  - "Facilitator: NAME HERE"
  - "Recorder: NAME HERE"
  - "Resource Manager: NAME HERE"
date: "DATE HERE"
format: pdf
editor: source
execute:
  echo: true
  error: true
  message: false
  warning: false
bibliography: references.bib
---

```{r setup}
#| include: false
library(tidyverse)
library(mosaic)
library(broom)
library(emmeans)
```

In this lab you will utilize a published dataset from @dekort2020 to assess the experimental methods used in the study. The study seeks to understand plasticity in the phenotype (change in the appearance or function) of wild strawberry (Fragaria vesca) due to environmental conditions. The authors measured many different plant traits and environmental variables as part of their study and used a complex regression approach to understand the relationship between the variables. In order to conduct such a study, the authors needed to ensure equivalency across their treatment groups and plots, ultimately measuring the initial growth of the plants as a proxy measurement for "maternal effects" like seed quality. While the authors report that they controlled for "maternal effects", they do not provide any evidence that "maternal effects" are similar across sites or treatments.

We will be focusing on two variables measured:

-   `Treatment`: designates the soil moisture (dry, normal, or wet) randomly applied to each strawberry plant.

-   `EarlyGrowth`: measures the above-ground biomass (grams) of the strawberry plant.

**Research Question:** Does the early growth of strawberries differ between soil moisture treatments?

```{r data}
strawberry_data <- read.csv("data/DeKort-data.csv")
head(strawberry_data)
```

1.  What is the observational unit?

2.  What are the explanatory and response variables of interest? What are their data types?

-   Explanatory:

-   Response:

4.  Write your parameters of interest *(Hint: there should be 3 of them)*.

5.  Create facetted histograms of the `EarlyGrowth` across the three soil moisture `Treatment`.

```{r}
ggplot(data = strawberry_data, 
       mapping = aes(x = ________)
       ) + 
  geom_histogram(binwidth = 0.25, color = "white") +
  facet_wrap(~ _______, ncol = 1) +
  labs(title = "_______________",
       x = "_______________",
       y = "_______________")
```

6.  Create side-by-side box plots of the `EarlyGrowth` across the three soil moisture `Treatment`.

```{r}
ggplot(data = strawberry_data, 
       mapping = aes(x = ________,
                     y = ________)
       ) + 
  geom_boxplot() +
  labs(title = "_______________",
       x = "_______________",
       y = "_______________")
```

7.  Use the `favstats()` function to obtain the summary of the observed data.

```{r}
favstats(______ ~ __________,
         data = strawberry_data)
```

8.  Using the visualizations and summary values of the observed data above, describe in your own words what you are observing in the data. For example, do you think there will be a difference in early-growth biomass between soil moisture treatments? Which soil moisture treatment has the highest early-growth biomass? The lowest? The largest spread? Etc.

9.  Write out the null and alternative hypotheses (in symbols), in the context of this study.

10. Check the assumptions for using ANOVA to investigate the research question.

-   Independence between groups:

-   Independence within groups:

-   Equal variance:

-   Sufficient Sample Size / Normality:

11. Fill in the code with these necessary pieces to conduct an ANOVA.

```{r}
strawberry_model <- aov(______ ~ ______,
                 data = _______)

_________ |> 
  tidy()

```

12. At an $\alpha = 0.05$ significance level, state a conclusion in context of the data. Make sure to include the F-statistic, appropriate degrees of freedom, and associated p-value.

13. Based on your answer, do you think it is appropriate for the authors to claim they controlled for "maternal effects"?
